Well-quasi-orderings on word languages

### 2024-10-17 src/introduction.tex
restructuring paper
Double knowledge

\AP When considering an algorithm based on a \kl{well-quasi-ordering}, the
runtime of the algorithm is deeply related to the ``complexity'' of the
underlying quasi-order \todo{CITE???}. One way to measure this complexity is to
consider its so-called \intro{ordinal invariants}: for instance, the
\intro{maximal order type} (or m.o.t.), originally defined by De Jongh and
Parikh \cite{dejongh77}, is the order type of the maximal linearization of a
well-quasi-ordered set. In the case of a finite set, the m.o.t. is precisely
the size of the set. Better runtime bounds were obtained by considering two
other parameters\todo{cite sylvain width/ HDR?}: the \intro{ordinal height}
introduced by Schmidt \cite{schmidt81}, and the \intro{ordinal width} of Kříž
and Thomas \cite{kriz90b}. Therefore, when characterizing
\kl{well-quasi-ordered} languages, we will also be interested in computing
bounds on their \kl{ordinal invariants}. We refer to the
\cref{ordinal-invariants:subsec} for a more detailed introduction to these
parameters and ordinal computations in general. For the moment, let us just
remark that all languages $L \subseteq \Sigma^*$ have an \kl{ordinal height} of
at most $\omega$.

@@

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{prefix relation} if and only if it is a finite union of \kl{chains} of the \kl{prefix relation}.
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{suffix relation} if and only if it is a finite union of \kl{chains} of the \kl{suffix relation}.
\end{enumerate}
}

Note that \kl{chains} are the simplest possible \kl{well-quasi-ordered} sets
(they are totally ordered and well-founded) and it is known that finite unions
of \kl{well-quasi-ordered} sets are \kl{well-quasi-ordered}. As a consequence,
the above characterization states that only the simplest possible
\kl{well-quasi-ordered} sets are \kl{well-quasi-ordered} by the \kl{prefix
relation}. Let us furthermore highlight that this characterization holds
without any restriction on the decidability of the language $L$ itself, but
heavily relies on the assumption that $\Sigma$ is finite. This allows us to
derive the following bounds on the \kl{ordinal invariants} of such
\kl{well-quasi-ordered} languages, which can be interpreted in two dual ways:
first, these languages are extremely simple, which means that one could have
hoped for using directly another well-quasi-ordered set without resorting to
finite words; second, any time one encounters a \kl{well-quasi-ordered}
language, it is going to provide relatively efficient algorithms.

\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{2}
    \item All languages $L \subseteq \Sigma^*$ are 
        \kl{well-quasi-ordered} by the \kl{prefix relation} 
        (resp. \kl{suffix relation}) have an \kl{ordinal height} of at most $\omega$,
        a finite \kl{ordinal width}, and a \kl{maximal order type} strictly
        below $\omega^2$.
\end{enumerate}
}

@@

\AP
Finally, we provide a decision procedure for checking whether a language is
\kl{well-quasi-ordered}, which takes as inputs languages recognized
by \kl{amalgamation systems} \cite{ASZZ24}. Such systems will be
formally introduced in \cref{amalgamation-systems:subsec}, but for the moment
let us just say that they include many classical computational models such as
finite automata, context-free grammars, and petri nets \cite{ASZZ24}. 
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{5}
    \item There is a decision procedure for checking whether a language $L$ recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation}, or the \kl{infix relation}. Furthermore, if it is \kl{well-quasi-ordered}, then $L$ is a \kl{bounded language}.
\end{enumerate}
}
Notice that the above result implies that the hypothesis of
\kl{bounded languages} on the theoretical side is not a restriction at all in
practice.

@@

\subparagraph{Outline} In
\cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} and prove in
\cref{prefixes:thm} the characterization of \kl{well-quasi-ordered}
languages by the \kl{prefix relation}. In
\cref{infixes-regular:sec}, we
obtain the \kl[infix relation]{infix} analogue of \cref{prefixes:thm}
specifically for regular languages
(\cref{infix-finite-automata:thm}). Finally, we generalize this result to all
\kl{amalgamation systems} in \cref{infixes-amalgamation:sec}
in
(\cref{infix-amalgamation:thm}).

>>>>

\AP When considering an algorithm based on a \kl{well-quasi-ordering}, the
runtime of the algorithm is deeply related to the ``complexity'' of the
underlying quasi-order \todo{CITE???}. One way to measure this complexity is to
consider its so-called \kl{ordinal invariants}: for instance, the
\kl{maximal order type} (or m.o.t.), originally defined by De Jongh and
Parikh \cite{dejongh77}, is the order type of the maximal linearization of a
well-quasi-ordered set. In the case of a finite set, the m.o.t. is precisely
the size of the set. Better runtime bounds were obtained by considering two
other parameters\todo{cite sylvain width/ HDR?}: the \kl{ordinal height}
introduced by Schmidt \cite{schmidt81}, and the \kl{ordinal width} of Kříž
and Thomas \cite{kriz90b}. Therefore, when characterizing
\kl{well-quasi-ordered} languages, we will also be interested in computing
bounds on their \kl{ordinal invariants}. We refer to the
\cref{ordinal-invariants:subsec} for a more detailed introduction to these
parameters and ordinal computations in general. For the moment, let us just
remark that all languages $L \subseteq \Sigma^*$ have an \kl{ordinal height} of
at most $\omega$.

@@

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{prefix relation} if and only if it is a finite union of \kl{chains} of the \kl{prefix relation}.
    \hfill (\cref{prefixes:thm} p.\pageref{prefixes:thm})
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{suffix relation} if and only if it is a finite union of \kl{chains} of the \kl{suffix relation}.
    \hfill (\cref{prefixes:thm} p.\pageref{prefixes:thm})
\end{enumerate}
}

Note that \kl{chains} are the simplest possible \kl{well-quasi-ordered} sets
(they are totally ordered and well-founded) and it is known that finite unions
of \kl{well-quasi-ordered} sets are \kl{well-quasi-ordered}. As a consequence,
the above characterization states that only the simplest possible
\kl{well-quasi-ordered} sets are \kl{well-quasi-ordered} by the \kl{prefix
relation}. Let us furthermore highlight that this characterization holds
without any restriction on the decidability of the language $L$ itself, but
heavily relies on the assumption that $\Sigma$ is finite. This allows us to
derive the following bounds on the \kl{ordinal invariants} of such
\kl{well-quasi-ordered} languages, which can be interpreted in two dual ways:
first, these languages are extremely simple, which means that one could have
hoped for using directly another well-quasi-ordered set without resorting to
finite words; second, any time one encounters a \kl{well-quasi-ordered}
language, it is going to provide relatively efficient algorithms.
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{2}
    \item All languages $L \subseteq \Sigma^*$ are 
        \kl{well-quasi-ordered} by the \kl{prefix relation} 
        (resp. \kl{suffix relation}) have an \kl{ordinal height} of at most $\omega$,
        a finite \kl{ordinal width}, and a \kl{maximal order type} strictly
        below $\omega^2$.
        \hfill (\cref{prefixes-ordinal-invariants:cor} p.\pageref{prefixes-ordinal-invariants:cor})
\end{enumerate}
}

@@

\AP
Finally, we provide a decision procedure for checking whether a language is
\kl{well-quasi-ordered}, which takes as inputs languages recognized
by \kl{amalgamation systems} \cite{ASZZ24}. Such systems will be
formally introduced in \cref{amalgamation-systems:subsec}, but for the moment
let us just say that they include many classical computational models such as
finite automata, context-free grammars, and petri nets \cite{ASZZ24}. 
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{5}
    \item There is a decision procedure for checking whether a language $L$ 
        recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered}
        by the \kl{prefix relation}, the \kl{suffix relation},
        or the \kl{infix relation}. Furthermore, if it is \kl{well-quasi-ordered}, 
        then $L$ is a \kl{bounded language}.
        \hfill (\cref{infix-wqo-is-emptiness:thm} p.\pageref{infix-wqo-is-emptiness:thm})
\end{enumerate}
}
Notice that the above result implies that the hypothesis of
\kl{bounded languages} on the theoretical side is not a restriction at all in
practice.

@@

\subparagraph{Outline} We introduce in \cref{prelims:sec} the
necessary background on \kl{well-quasi-orders} and \kl{ordinal invariants}.
In
\cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} and prove in
\cref{prefixes:thm} the characterization of \kl{well-quasi-ordered}
languages by the \kl{prefix relation}. In
\cref{infixes-regular:sec}, we
obtain the \kl[infix relation]{infix} analogue of \cref{prefixes:thm}
specifically for regular languages
(\cref{infix-finite-automata:thm}). Finally, we generalize this result to all
\kl{amalgamation systems} in \cref{infixes-amalgamation:sec}
in
(\cref{infix-amalgamation:thm}).
